Fractional Types

### 2013-02-12 frac.tex
several comments

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! foldB = fold_B
%! unfoldB = unfold_B
%! foldN = fold_N
%! unfoldN = unfold_N
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! :-* = \multimap
%! :-+ = \multimap^{+}
%! emptyset = \emptyset

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}

@@

@@

@@

\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \} \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

@@

@@

\begin{verbatim}
In the rules above it is unclear as to which v is chosen in the rule
 for eta. It is unclear that these rules are logically reversible.
\end{verbatim}

@@

\subsection{Relational Semantics}

@@

\noindent
We now have primitives:
%subcode{opsem} include main
% swap* '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% swap+ '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) ~|~ v1 in b1, v2 in b2]}
% assocl* '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) ~|~ v1 in b1, v2 in b2, v3 in b3]}
etc..

@@

We can define evaluation as follows: 
%
{{ c v = {[ v' ~|~ (v, v') in c ]} }}.

However this is unpleasant since it is unclear how to evaluate if
{{c~v}} and {{c^{dagger} v'}} are dual by evaluation. This ties into a
deeper issue of interpreting relations as programs.

Q) If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
are sets {{b}} and the morphisms are the relations between objects of
equal size.

>>>>

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
\newcommand{\roshan}[1]{\textsc{Roshan says:} 
  \textit{#1}
}

@@

%subcode-inline{bnf-inline} name Pi
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! foldB = fold_B
%! unfoldB = unfold_B
%! foldN = fold_N
%! unfoldN = unfold_N
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! :-* = \multimap
%! :-+ = \multimap^{+}
%! emptyset = \emptyset

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

@@

\roshan{It is unclear what the Id types refer to. We should provide
  some more context here}

@@

\roshan{It is unclear how fractions help differentiate between inputs
  and outputs. On first reading this seemed promising, but nothing
  in the rest of the paper substantiates this point of view.}

\roshan{I think we should arrive at relational programming as a means
  of giving semantics to fractionals, rather than the other way
  round. To show that relational programming is reversible, one does
  not need to resort to fractionals. Everyone understand that
  relations can be reversed and none of this setup with {{Pi}} is
  required for that. Our key point should be to communicate that
  fractionals have a clean semantics in the setting of relational
  programming rather than the other way round.}

@@

\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
%subcode{opsem} include main
%! <- = \leftarrow
%! union = \cup
% [[1]] '= {[ () ]}
% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
\end{definition}

@@

\roshan{In the judgments below it is misleading that only {{eta*}}
  appears indexed by a type. To construct each family of relations,
  the corresponding definition must be indexed by their types. In
  other words, there is nothing special about {{eta*}}. Further, I am
  unsure of how to format this so that the combinators are correctly
  indexed by their types for both primitives and for composition
  combinators. We could do the following, but it seems to be a
  overkill.}

%subcode{proof} include main
%@ v1 in [[b1]]
%@ v2 in [[b2]]
%@ v3 in [[b3]]
%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)

@@

\roshan{ In the rules above it is unclear as to which v is chosen in
  the rule for eta. It is unclear that these rules are logically
  reversible.  }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relational Semantics}

@@

\noindent
We now have primitives:
%subcode{opsem} include main
% swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
%                     && ~~~| v1 in b1, v2 in b2]}
% assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
%                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
etc..

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation}

I think defining eval is an open problem.  Consider a few relations:

\begin{enumerate}
\item 
{{empty = {[ ]} }}

\item 
{{full =  {[ R, G, B ]} * {[ R, G, B ]} }}

\item 
{{id  =  {[ (R, R), (G, G), (B, B) ]} }}

\item 
{{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
\end{enumerate}

Can we define an {{eval}} that meaningfully distinguishes these? What
can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
several possible {{eval}} definitions:

\begin{enumerate}
\item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
  eval maps {{b1 -> b2}}.

\item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.

\item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
  {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.

\end{enumerate}

For the relations above, none of the suggested eval definitions give a
logically reversible semantics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \subsection{Model}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Completeness}

\roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
are sets {{b}} and the morphisms are the relations between objects of
equal size.}

\roshan{ I suspect that all relations can be constructed, but we have
  to show this. This is not obvious.}
